---
title: "One Sample Inference"
subtitle: "Stats 7"
author: "Mary Ryan"
date: "Sept. XX, 2020"
institute: "University of California, Irvine"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "chocolate-fonts", "headerPosition.css"]#like lucy
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      slideNumberFormat: "%current%"
    seal: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.align='center', out.width="53%", warning=FALSE)
```

```{r load libraries, include=FALSE, comment='#'}
#install.packages( 'tidyverse' )
#install.packages( 'readxl' )

library( tidyverse, quietly=T, warn.conflicts = F )
library( knitr, quietly=T, warn.conflicts = F )
library( readxl, quietly=T, warn.conflicts = F )
#library( tufte, quietly=T, warn.conflicts = F )
library( quantmod, quietly=T, warn.conflicts = F )

```

```{r reduce code spacing, include = FALSE}
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
```

```{r load data, include=F}

#avocado <- read_excel( "avocado.xlsx" )

```

layout: true
class: 

<!-- Old footer font color: #00A895 -->
<!-- old footer background color: #383838 -->

<!-- footer -->
 <div style="position:fixed; bottom:10px; left:4px; font-size: 12pt; color: #2DD8A5; background-color: #545454; width:93.5%">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Mary Ryan</div> <!--&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; -->
<div style="position:fixed; bottom:10px; left:500px; font-size: 12pt; color: #2DD8A5">One Sample Inference</div> <!--&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; -->
<div style="position:fixed; bottom:10px; right:92px; font-size: 12pt; color: #2DD8A5">Stats 7</div>

---

<div style="position:fixed; left:50px; right:50px; top:125px; padding:30px; margin:auto; background-color:#C0F3E4; border-radius:15px">
<p>This class is being conducted over Zoom. As the instructor, I will be .alert[recording] this session. I have disabled the recording feature for others so that no one else will be able to record this session. I will be posting this session to the course’s website.</p>

<p>If you have privacy concerns and .alert[do not wish to appear in the recording], you may turn video off (click .alert[“stop video”]) so that Zoom does not record you.</p>

<p>The chat box is always open for discussion and questions to the entire class. You may also send messages privately to the instructor or the TAs. Please note that Zoom saves all chat transcripts.</p>

<p>I create a live transcription of each session using <a href="https://otter.ai/">Otter.ai</a>. This means that Otter.ai will transcribe anything spoken over the Zoom audio. The transcript will be posted with the session video on the course website.</p>
</div>

---

class: title-slide2

# <center> One Sample Inference </center>
## <center> Stats 7 </center>
### <center> Mary Ryan </center>
### .center[Aug. 27, 2020]

<!-- social media info -->
<div style="position:fixed; bottom:40px; left:70px;">
<p><img src="https://du11hjcvx0uqb.cloudfront.net/br/dist/images/favicon-e10d657a73.ico" width="35"/> Course website:</p>
<p><a href="https://canvas.eee.uci.edu/courses/28451"> https://canvas.eee.uci.edu/courses/28451 </a></p>

<p><img src="https://svgsilh.com/svg/1873373.svg" width="35"/> Slides can be found at:</p>
</p><a href="https://maryryan.github.io/stats7-SS2-2020-slides/stats7-SS2-2020-samplingBias/stats7-SS2-2020-samplingBias"> https://maryryan.github.io/stats7-SS2-2020-slides/stats7-SS2-2020-samplingBias/stats7-SS2-2020-samplingBias </a></p>
</div>



---

# How to Perfom a Hypothesis Test

1. State your hypotheses

2. Calculate your Z-statistic

3. Find your Z-critical ( $Z^*$ ) value

4. Compare your Z-statistic to $Z^*$
  
5. Draw a conclusion

6. Interpret you conclusion in context
---

# Step 1: State your hypotheses

.alert[Question of Interest Here]

- **Null hypothesis** ( $H_0$ ) represents "status quo", or what we assume to be true without more evidence
- Question will usually give you the **alternative hypothesis** ( $H_A$ )
- Two types of tests:
   - Two-sided test ( $H_A: p \ne p_0$ )
   - One-sided test ( $H_A: p > p_0$, or $H_A: p < p_0$ )
   
- Always write hypotheses in terms of the parameter of interest (not the sample estimate)

---

# Step 2: Calculate your Z-statistic

$$Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$$

---

# Step 3: Find $Z^*$

- For two-sided tests, we split our **error rate** into both the tails
```{r twoSided, echo=FALSE, out.width=500, dpi=500}
par(mar = c(5.1, 0.5, 4.1, 0.5))
x <- seq(-6,6, length=1000)
y <- dnorm(x)
plot(x, y, type="l", lwd=1,
     xaxt="n", xlab="",
     yaxt="n", ylab="")

abline(v=1.96)
abline(v=-1.96)

text(c(-1.96,1.96), rep(-0.04,2), c(expression(-Z^"*"), expression(Z^"*")), xpd=NA)

text(0, 0.1, "Not crazy\n values", col="forest green")
text(c(-3,3), 0.1, "Woah!", col="red")
text(c(-2.5, 2.5), 0.0, expression(alpha/2), cex=0.9)

```

---

# Step 3: Find $Z^*$

- For one-sided tests, we can keep all of the error rate in the tail we're interested in
```{r oneSided, echo=FALSE, out.width=450, dpi=500}
par(mfrow=c(1,2), mar = c(5.1, 0.5, 4.1, 0.5))

# greater than plot
plot(x, y, type="l", lwd=1,
     xaxt="n", xlab="",
     yaxt="n", ylab="")

abline(v=1.96)
#abline(v=-1.96)

text(1.96, -0.04, expression(Z^"*"), xpd=NA)

text(0, 0.1, "Not crazy\n values", col="forest green")
text(3, 0.1, "Woah!", col="red")
text(2.5, 0.0, expression(alpha), cex=0.9)

text(0, -0.08, expression(H[A]: p > p[0]), xpd=NA, cex=1.2)

# less than plot 
plot(x, y, type="l", lwd=1,
     xaxt="n", xlab="",
     yaxt="n", ylab="")

#abline(v=1.96)
abline(v=-1.96)

text(-1.96, -0.04, expression(-Z^"*"), xpd=NA)

text(0, 0.1, "Not crazy\n values", col="forest green")
text(-3, 0.1, "Woah!", col="red")
text(-2.5, 0.0, expression(alpha), cex=0.9)

text(0, -0.08, expression(H[A]: p < p[0]), xpd=NA, cex=1.2)


```

---

# Steps 4, 5, 6

---

# Step 5: Draw a conclusion

- For two-sided test:
   - If |Z| > $Z^*$, __reject__ the null
   - If |Z| < $Z^*$, __fail to reject__ the null
   
- For one-sided test:
   - Greater-than test:
      - If Z > $Z^*$, reject the null
      - If Z < $Z^*$, fail to reject the null
      
   - Less-than test:
      - If Z < $-Z^*$, reject the null
      - If Z > $-Z^*$, fail to reject the null

---

# Step 6: Interpret your conclusion in context

- For a rejection:

- For a fail-to-reject:
